Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Made numbers fit with unit trickery.
Multiple character variables wrapped in mathit.

\begin{figure}
\begin{gather*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep&
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
\multicolumn{3}{r@{}}{
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)}
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this and the following section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each optimization. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all subsequent stores.

@@

\begin{align*}
force &: \Store \times \Addr \to \Set(\Value) \\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\} \\
ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\begin{align*}
replay &: \StoreDelta \times \Store \to \Store \\
replay(\mtlst, \msto) &= \msto \\
replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', replay(\msdiff,\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

Here $concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
concat(\varnothing) &= \mtlst \\
concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
\end{align*}

@@

\begin{figure}
\begin{align*}
\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
merge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

@@

\begin{figure}
\include{bench-overview}
\caption{Overview performance comparison between baseline and
  optimized analyzer (entries of \text{{\small $t$}} mean timeout, and \text{{\small $m$}} mean out of memory).}
\label{fig:bench-overview}
\end{figure}

>>>>

\begin{figure}
\begin{gather*}
\begin{align*}
\eval(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep&
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
\multicolumn{3}{r@{}}{
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)}
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

The uniform $k$-CFA allocation strategy would make $\eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this and the following section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each optimization. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c, \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all subsequent stores.

@@

\begin{align*}
force &: \Store \times \Addr \to \Set(\Value) \\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\} \\
ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

\begin{align*}
\replay &: \StoreDelta \times \Store \to \Store \\
\replay(\mtlst, \msto) &= \msto \\
\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

\begin{align*}
(\states, \msto) &\mathrel{\widehat{\machstep}} (\states\cup \states', \replay(\msdiff,\msto)) \\
\mbox{ where } (\states', \msdiff) &= \step^*(\varnothing, \states, \mtlst) \\
\step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
\step^*(S, \{c\}\cup \states, \msdiff &= \step^*(S\cup cs^*, \states, \msdiff^*) \\
\states^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= \concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
\concat(\varnothing) &= \mtlst \\
\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
\end{align*}

@@

\begin{figure}
\begin{align*}
\interpdelta(\msto,\ {\tt qlist}) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto,\ {\tt qlist}, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= \merge(v, \bigsqcup(vs ...)) \\
\merge(v, v) &= v \\
\merge(n, m) &= {\tt Number} \\
\merge(n, v) &= {\tt QData} \\
\merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
\merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
\merge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

@@

\begin{figure}
\centering
\include{bench-overview}
\caption{Overview performance comparison between baseline and
  optimized analyzer (entries of \text{{\small $t$}} mean timeout, and \text{{\small $m$}} mean out of memory).}
\label{fig:bench-overview}
\end{figure}
